---
title: Sharing Trained Models via Hugging Face Hub
subtitle: A workflow for fitting, shrinking, uploading, and retrieving model objects
---

```{r}
#| label: setup-python
#| include: false

reticulate::use_python(Sys.getenv("RETICULATE_PYTHON"), required = TRUE)
```

## Motivation

When collaborating on empirical research or peer-reviewing assignments, it is often useful to share fitted model objects. This allows a reviewer to load your model and verify predictions without re-running potentially expensive training code.

This post walks through a minimal end-to-end workflow:

1. Fit a model
2. Reduce its size
3. Upload it to Hugging Face Hub
4. Download and generate predictions from a fresh session

## Setup

::: {.panel-tabset group="language"}

## R

We load six packages. `tidyfinance` provides access to financial data. `tidymodels` handles the full modelling workflow. `butcher` trims fitted model objects. `httr2` sends HTTP requests to the Hugging Face API. `readr` reads and writes `.rds` files. `base64enc` encodes binary files as base64 text for the upload API.

```{r}
#| label: load-packages-r

library(tidyfinance)
library(tidymodels)
library(butcher)
library(httr2)
library(readr)
library(base64enc)
```

## Python

We import six libraries. `tidyfinance` provides the same financial data as the R version. `pandas` and `numpy` handle data manipulation. `sklearn` covers preprocessing, modelling, and cross-validation. `joblib` serialises model objects to disk. `huggingface_hub` uploads and downloads files from Hugging Face.

```{python}
#| label: load-packages-py

import tidyfinance as tf
import pandas as pd
import numpy as np
from sklearn.linear_model import Ridge
from sklearn.preprocessing import StandardScaler
from sklearn.pipeline import Pipeline
from sklearn.model_selection import TimeSeriesSplit, GridSearchCV
from sklearn.metrics import root_mean_squared_error, r2_score, mean_absolute_error
import joblib
from huggingface_hub import HfApi, hf_hub_download
```

:::

## A simple financial application

Suppose we want to predict monthly S&P 500 excess returns using macroeconomic predictors. The target variable is `rp_div`, the excess return on the S&P 500. The dataset also contains a set of macro predictors, all already lagged.

::: {.panel-tabset group="language"}

## R

`download_data()` fetches monthly macro predictors directly into a tibble.

```{r}
#| label: download-data-r

data <- download_data("macro_predictors", "monthly")
```

## Python

`tf.download_data()` returns the same dataset as a pandas DataFrame.

```{python}
#| label: download-data-py

data = tf.download_data(
    domain="macro_predictors",
    dataset="monthly",
)
```

:::

## Fit a model

### Split the data

::: {.panel-tabset group="language"}

## R

Because the data are a time series, we use `initial_time_split()` rather than a random split. This puts the first 80% of observations (sorted by date) into the training set and the remaining 20% into the test set. A random split would let future observations leak into training, which would give an overly optimistic picture of model performance.

```{r}
#| label: train-test-split-r

split      <- initial_time_split(data, prop = 0.8)
train_data <- training(split)
test_data  <- testing(split)
```

## Python

We sort by `date` first to guarantee chronological order, then slice at the 80% mark. We also separate the feature matrix `X` from the target vector `y` for both splits, dropping `date` from the features since it carries no predictive signal.

```{python}
#| label: train-test-split-py

data = data.sort_values("date").reset_index(drop=True)

split_idx  = int(len(data) * 0.8)
train_data = data.iloc[:split_idx]
test_data  = data.iloc[split_idx:]

feature_cols = [c for c in data.columns if c not in ["date", "rp_div"]]

X_train = train_data[feature_cols]
y_train = train_data["rp_div"]
X_test  = test_data[feature_cols]
y_test  = test_data["rp_div"]
```

:::

### Preprocess

::: {.panel-tabset group="language"}

## R

A recipe defines the preprocessing steps applied to the data before modelling. We tell it to predict `rp_div` from all other columns, then remove `date`, and then standardise each predictor to zero mean and unit variance. Standardisation matters here because ridge regression shrinks coefficients towards zero and the amount of shrinkage depends on the scale of each variable.

```{r}
#| label: define-recipe-r

rec <- recipe(rp_div ~ ., data = train_data) |>
  step_rm(date) |>
  step_normalize(all_predictors())
```

## Python

In sklearn, preprocessing and modelling combine into a `Pipeline`. We chain a `StandardScaler` (which standardises to zero mean and unit variance) with a `Ridge` model. The pipeline applies the scaler to training data and carries those parameters forward when transforming new data.

```{python}
#| label: define-recipe-py

pipe = Pipeline([
    ("scaler", StandardScaler()),
    ("ridge",  Ridge())
])
```

:::

### Tune and fit

::: {.panel-tabset group="language"}

## R

We fix `mixture = 0`, which gives a pure ridge regression, and tune only the penalty parameter via cross-validation. Because the data are a time series, we use `rolling_origin()` to create folds that always train on past observations and validate on future ones, preserving temporal order throughout the tuning process. Each fold uses an initial window of 120 months for training and assesses on the following 60 months, with `cumulative = FALSE` so the training window does not grow as more data become available. Setting `skip = 59` advances the window by 60 months between folds rather than one month at a time, which reduces the total number of folds from several hundred to around twelve and keeps computation tractable.

`tune_grid()` fits the model for each of 30 candidate penalty values on every fold and records RMSE, traditional R-squared, and MAE. `select_best()` picks the penalty with the lowest average RMSE across folds. `finalize_workflow()` plugs that value back into the workflow, and `fit()` trains the final model on the full training set.

```{r}
#| label: tune-model-r

enet_spec <- linear_reg(penalty = tune(), mixture = 0) |>
  set_engine("glmnet")

enet_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(enet_spec)

rolling_folds <- rolling_origin(
  train_data,
  initial    = 120,
  assess     = 60,
  cumulative = FALSE,
  skip       = 59
)

tuned <- tune_grid(
  enet_wf,
  resamples = rolling_folds,
  grid      = 30,
  metrics   = metric_set(rmse, rsq_trad, mae)
)

enet_wf_final <- enet_wf |>
  finalize_workflow(select_best(tuned, metric = "rmse")) |>
  fit(train_data)
```

## Python

`TimeSeriesSplit` creates 12 folds with an expanding training window, so each fold always trains on earlier data and validates on later data. This is the sklearn equivalent of `rolling_origin()` with `cumulative = TRUE`. We search over 30 candidate values of `alpha` (the ridge penalty) on a log scale. `GridSearchCV` fits the full pipeline (scaler plus ridge model) for each combination and selects the value with the best cross-validated RMSE. The best estimator is already fitted on the last fold's training data; we refit it on the full training set to use all available information.

```{python}
#| label: tune-model-py

tscv = TimeSeriesSplit(n_splits=12)

param_grid = {"ridge__alpha": np.logspace(-4, 0, 30)}

search = GridSearchCV(
    pipe,
    param_grid,
    cv      = tscv,
    scoring = "neg_root_mean_squared_error",
    refit   = True
)
search.fit(X_train, y_train)

best_model = search.best_estimator_
```

:::

## Reduce model size

::: {.panel-tabset group="language"}

## R

A fitted workflow in R carries a lot of extra baggage: the training data, the call environment, and other objects that are only needed during fitting. `butcher()` strips those away. The result is a much smaller file that still produces identical predictions.

```{r}
#| label: butcher-and-save-r

enet_wf_final |> butcher() |> write_rds("model_ridge.rds")
```

## Python

sklearn pipelines are already compact. `joblib.dump()` serialises the fitted pipeline to disk. `joblib` is preferred over `pickle` for sklearn objects because it handles large numpy arrays more efficiently.

```{python}
#| label: save-model-py

joblib.dump(best_model, "model_ridge.pkl")
```

:::

## Upload to Hugging Face Hub

You need a [Hugging Face access token](https://huggingface.co/settings/tokens) with write permissions. Store it as an environment variable so that it is not exposed in your code.

::: {.panel-tabset group="language"}

## R

You can add `HF_TOKEN=hf_your_token_here` to your `.Renviron` file (run `usethis::edit_r_environ()` to open it) so the token is available across sessions.

```{r}
#| label: set-hf-token-r
#| eval: false

Sys.setenv(HF_TOKEN = "hf_your_token_here")
```

## Python

Set the token as an environment variable before running the upload code. The `huggingface_hub` library reads it automatically via `HfApi()`.

```{python}
#| label: set-hf-token-py
#| eval: false

import os
os.environ["HF_TOKEN"] = "hf_your_token_here"
```

:::

### Step 1: Create a repository

::: {.panel-tabset group="language"}

## R

We read the token from the environment and send a POST request to the Hugging Face API. This creates a new public model repository under your account.

```{r}
#| label: hf-create-repo-r

hf_token <- Sys.getenv("HF_TOKEN")

request("https://huggingface.co/api/repos/create") |>
  req_auth_bearer_token(hf_token) |>
  req_body_json(list(
    name = "ma2-model-1",
    type = "model",
    private = FALSE
  )) |>
  req_perform()
```

## Python

`HfApi.create_repo()` creates a new public model repository under your account in a single call.

```{python}
#| label: hf-create-repo-py

api = HfApi()
api.create_repo(repo_id="your-username/ma2-model-1", repo_type="model", private=False)
```

:::

### Step 2: Upload the model file

::: {.panel-tabset group="language"}

## R

Hugging Face stores files via a commit API. We read the `.rds` file as raw bytes, encode it to base64 (a text-safe representation of binary data), and send it as the body of a commit request. Replace `your-username` with your Hugging Face username.

```{r}
#| eval: false
#| label: hf-username-r

hf_username <- "your-username"
```

```{r}
#| echo: false
hf_username <- "voigtstefan"
```

```{r}
#| label: hf-upload-model-r

repo_name <- "ma2-model-1"
file_name <- "model_ridge.rds"

model_bytes  <- readBin(file_name, "raw", file.size(file_name))
model_base64 <- base64enc::base64encode(model_bytes)

request(
  paste0("https://huggingface.co/api/models/", hf_username, "/", repo_name, "/commit/main")
) |>
  req_auth_bearer_token(hf_token) |>
  req_body_json(list(
    summary = "Upload ridge model",
    files = list(
      list(
        path = file_name,
        content = model_base64,
        encoding = "base64"
      )
    )
  )) |>
  req_perform()
```

## Python

`upload_file()` handles the commit workflow internally. We point it at the local file and specify where to put it in the repository. Replace `your-username` with your Hugging Face username.

```{python}
#| label: hf-upload-model-py

api.upload_file(
    path_or_fileobj = "model_ridge.pkl",
    path_in_repo    = "model_ridge.pkl",
    repo_id         = "your-username/ma2-model-1",
    repo_type       = "model"
)
```

:::

After this step, your model is publicly available at:
`https://huggingface.co/<your-username>/ma2-model-1`

## Download and predict from a fresh session

A peer reviewer only needs your Hugging Face username.

::: {.panel-tabset group="language"}

## R

`download.file()` saves the file to disk, `read_rds()` loads it back into R, and `augment()` attaches the model predictions to the test data as a new column `.pred`. The `metrics()` call then computes RMSE, traditional R-squared, and MAE against the true values.

```{r}
#| label: download-and-predict-r

model_url <- paste0("https://huggingface.co/", hf_username, "/ma2-model-1/resolve/main/model_ridge.rds")
download.file(model_url, destfile = "model_ridge.rds", mode = "wb")
loaded_model <- read_rds("model_ridge.rds")

augment(loaded_model, test_data) |>
  metrics(truth = rp_div, estimate = .pred)
```

## Python

`hf_hub_download()` fetches the file and returns its local path. We load it with `joblib`, generate predictions with `predict()`, and compute the same three metrics as in R.

```{python}
#| label: download-and-predict-py

model_path   = hf_hub_download(repo_id="your-username/ma2-model-1", filename="model_ridge.pkl")
loaded_model = joblib.load(model_path)

y_pred = loaded_model.predict(X_test)

print(f"RMSE: {root_mean_squared_error(y_test, y_pred):.4f}")
print(f"R2:   {r2_score(y_test, y_pred):.4f}")
print(f"MAE:  {mean_absolute_error(y_test, y_pred):.4f}")
```

:::

## Summary

::: {.panel-tabset group="language"}

## R

| Step | Function | Package |
|------|----------|---------|
| Fit model | `fit()` | `tidymodels` |
| Reduce size | `butcher()` | `butcher` |
| Save to disk | `write_rds()` | `readr` |
| Upload to HF Hub | `req_perform()` | `httr2` |
| Download | `download.file()` | base R |
| Load and predict | `read_rds()` + `augment()` | `readr` / `tidymodels` |

## Python

| Step | Function | Package |
|------|----------|---------|
| Fit model | `fit()` | `sklearn` |
| Save to disk | `joblib.dump()` | `joblib` |
| Upload to HF Hub | `upload_file()` | `huggingface_hub` |
| Download | `hf_hub_download()` | `huggingface_hub` |
| Load and predict | `joblib.load()` + `predict()` | `joblib` / `sklearn` |

:::
