---
title: High-frequency econometrics
execute: 
  eval: true
---

Two recently developed packages simplify programmatic access to [LOBSTER](https://lobsterdata.com) data directly from your analysis environment:

- **R — `lobsteR`**: Available on [CRAN](https://cran.r-project.org/web/packages/lobsteR/index.html), `lobsteR` handles authentication, data requests, and downloads from the LOBSTER platform and integrates naturally with the `highfrequency` package for downstream analysis.
- **Python — `lobsterdata`**: A lightweight [Python client](https://github.com/rhuang10/lobsterdata) that lets you submit data construction requests, monitor their status, and download the resulting CSV files, covering the full asynchronous workflow of the LOBSTER API.

KU students can obtain access to LOBSTER data by writing to [stefan.voigt@econ.ku.dk](mailto:stefan.voigt@econ.ku.dk).

## Read in and process Lobster files

This exercise requires you mainly to familiarize yourself with [LOBSTER](https://lobsterdata.com), an online limit order book data tool that provides easy-to-use, high-quality data for the entire universe of NASDAQ traded stocks. I requested data based on their online interface and stored it on Absalon before running the code below.
Lobster files contain **all** trading messages (order submissions, cancellations, trades) for a pre-specified ticker that went through NASDAQ. For example, the sample file in Absalon contains the entire order book history for the ticker `SAP` until level 5 in January 2022.
LOBSTER generates a 'message' and an 'orderbook' file for each active trading day of a selected ticker. The 'orderbook' file contains the evolution of the limit order book up to the requested levels. The 'message' file contains indicators for the type of event causing an update of the limit order book in the requested price range. All events are time-stamped in seconds after midnight, with decimal precision of at least milliseconds and up to nanoseconds depending on the requested period.

The 'message' and 'order book' files are provided in the .csv format and can easily be read with any statistical software package.

**Exercises:**

- Familiarize yourself with the output documentation of Lobster files [on their homepage](https://lobsterdata.com/info/DataStructure.php). Download the file `SAP_2022-01-01_2022-01-31_5.7z` from Absalon and read them using `read_csv`. Lobster files always have the same naming convention, `ticker_date_34200000_57600000_filetype_level.csv`, where `filetype` denotes `message` or the corresponding `order book` snapshots. Take notes on what you think are essential cleaning steps to get a helpful dataset comprising order book and trading information.
- Merge the two files after you apply some intuitive naming convention such that you can work with one file which traces all information about messages and the corresponding order book snapshots.
Try to create a function `process_orderbook`, which performs essential cleaning of the data by making sure you filter out trading halts and aggregate order execution messages which may have triggered multiple limit orders.
More specifically, the function should complete the following generic steps: 

  1. Identify potential trading stops (consult the Lobster documentation for this) and discard all observations recorded during the window when trading was interrupted. 
  2. Sometimes, messages related to the daily opening and closing auction are included in the Lobster file (especially during days when trading stops earlier, e.g., before holidays). Remove such observations (opening auctions are identified as messages with `type == 6` and `ID == -1`, and closing auctions can be identified as messages with `type == 6` and `ID == -2`).
  3. Replace "empty" slots in the order book with `NA`s in case there was no liquidity quoted at higher order book levels. In that case, Lobster reports negative bid and large ask prices (above 199999). 
  4. Remove observations with seemingly crossed prices (when the best ask is below the best bid).
  5. Market orders are sometimes executed against multiple limit orders. In that case, Lobster records multiple transactions with an identical time stamp. Aggregate these transactions and generate the corresponding trade size and the (size-weighted) execution price. Also, retain the order book snapshot *after* the transaction has been executed. 
  6. The direction of a trade is hard to evaluate (not observable) if the transaction has been executed against a hidden order (`type == 5`). Create a proxy for the direction based on the executed price relative to the last observed order book snapshot.

You can use this function when you work with raw order book data from Lobster to produce a clean, analysis-ready order book. In the solution part, I provide some further explanations of my implementation.
- Use the function to clean the SAP Lobster output. 
- Summarize the order book by extracting the average midquote $q_t = (a_t + b_t)/2$ (where $a_t$ and $b_t$ denote the best bid and best ask), the average spread $S_t= (a_t - b_t)/q_t$ (in basis points relative to the concurrent midquote), the aggregate volume in USD and the quoted depth (in million USD 5 basis points from the concurrent best price on each side of the order book) for each 5-minute interval.  
- Provide a meaningful illustration of the order book dynamics during the particular trading day. **This is intentionally a very open question: There is so much information within an order book that it is impossible to highlight everything worth considering. I encourage you to post your visualization on the discussion board in Absalon.**

**Solutions:**

::: {.panel-tabset group="language"}
## R

```{r}
#| label: load-libraries-lobster
library(dplyr)
library(tidyr)
library(archive)
library(readr)
library(ggplot2)
library(gt)
library(lubridate)
library(forcats)
```

I use the following code to read in (general) Lobster files - change `ticker`, `date`, and `level` in the code snippet below.
```{r}
#| label: read-archive-files
files_available <- archive::archive("SAP_2022-01-01_2022-01-31_5.7z") |>
  mutate(ticker = gsub("(.*)_(.*)_342.*_(.*).csv", "\\1", path),
         date = gsub("(.*)_(.*)_342.*_(.*).csv", "\\2", path),
         level = gsub("(.*)_(.*)_342.*_(.*).csv", "\\3", path))

files_available
```

Lobster files always follow the same naming convention; thus, the code above can be recycled in a more general fashion for other Lobster files. Next, we read in data for one(!) of the `{r} nrow(files_available)/2` trading days. 

```{r}
#| label: fig-traded-prices
#| fig-cap: "Traded prices for SAP on January 3rd, 2022, based on execution messages (types 4 and 5)."
messages_raw <- read_csv(archive::archive_read("SAP_2022-01-01_2022-01-31_5.7z", 3),
col_names = c("ts", "type", "order_id", "m_size", "m_price", "direction", "null"),
col_type = cols(null = col_skip())) |>
mutate(ts = as.POSIXct(ts, origin = as.Date("2022-01-07"), tz ="GMT", format = "%Y-%m-%d %H:%M:%OS6"),
m_price = m_price / 10000) 

messages_raw |>
filter(type ==4| type ==5) |>
ggplot(aes(x = ts, y = m_price)) + geom_step() +
geom_point(data = messages_raw |> tail(1), color = "red", size = 3) +
labs(x = "", y = "USD", title = "Traded prices: SAP, January 3rd, 2022" )
```

Note my naming convention for column names. The command `col_type` makes sure that there is easy processing afterward.
By default, `ts` denotes the time in seconds since midnight (decimals are precise until the nanosecond level), and `price` always comes in 10.000 USD. `type` denotes the message type: `4`, for instance, corresponds to the execution of a visible order. The remaining variables are explained in more detail [here](https://lobsterdata.com/info/DataStructure.php).

Next, the corresponding orderbook snapshots contain `price` and quoted `size` for each of the `5` levels.

Each message is associated with the corresponding order book snapshot.
After merging the `message` and `orderbook` files, the entire data thus looks as follows.

The code below is rather flexible to handle Lobster files with more or less levels. Each column contains the corresponding prices and sizes on each side of the order book. More information is available on the documentation of Lobster. The output contains one file with all information available in the order book.
```{r}
#| label: read-orderbook
level <- 5
orderbook_raw <- read_csv(archive::archive_read("SAP_2022-01-01_2022-01-31_5.7z", 4),
                          col_names = paste(rep(c("ask_price", "ask_size", "bid_price", "bid_size"), level), rep(1:level, each = 4), sep = "_"),
                          cols(.default = col_double())) |>
  mutate_at(vars(contains("price")), ~./10000) 

orderbook <- bind_cols(messages_raw, orderbook_raw) 
orderbook
```

The order book contains `{r} nrow(orderbook)` observations. While the columns `ts` to `direction` help understand the specific messages, the remaining columns contain a snapshot of the order book at that particular time. 

Next, Lobster sometimes contains opening and closing auctions (in particular on days when trading hours deviate from the regular times). If you follow the documentation on Lobster, you will see that these times can be identified as I do in the code below. 

```{r}
#| label: identify-auctions
opening_auction <- orderbook |>
  filter((type == 6 & order_id == -1) | row_number() == 1) |>
  filter(row_number() == n()) |>
  pull(ts)
closing_auction <- orderbook |>
  filter((type == 6 & order_id == -2) | row_number() == n()) |>
  filter(row_number() == 1) |>
  pull(ts)

c(opening_auction, closing_auction)
```

In line with the exercise, the function `process_orderbook()` performs several essential steps. You can call it from within `R` with the following command.

```{r}
#| label: define-process-orderbook
process_orderbook <- function(orderbook) {
  # Did a trading halt happen? ----
  halt_index <- orderbook |>
    filter(type == 7 & direction == -1 & m_price == -1 / 10000 | type == 7 & direction == -1 & m_price == 1 / 10000)

  while (nrow(halt_index) > 1) {
    # Filter out messages that occurred in between trading halts
    cat("Trading halt detected")
    orderbook <- orderbook |>
      filter(ts < halt_index$ts[1] | ts > halt_index$ts[2])
    halt_index <- halt_index |> filter(row_number() > 2)
  }

  # Opening and closing auction ----
  # Discard everything before type 6 & ID -1 and everything after type 6 & ID -2

  opening_auction <- orderbook |>
    filter(
      type == 6,
      order_id == -1
    ) |>
    pull(ts)

  closing_auction <- orderbook |>
    filter(
      type == 6,
      order_id == -2
    ) |>
    pull(ts)

  if (length(opening_auction) != 1) {
    opening_auction <- orderbook |>
      select(ts) |>
      head(1) |>
      pull(ts) - lubridate::seconds(0.1)
  }
  if (length(closing_auction) != 1) {
    closing_auction <- orderbook |>
      select(ts) |>
      tail(1) |>
      pull(ts) + lubridate::seconds(0.1)
  }

  orderbook <- orderbook |> filter(ts > opening_auction & ts < closing_auction)
  orderbook <- orderbook |> filter(type != 6 & type != 7)

  # Replace "empty" slots in the order book (0 volume) with NA prices ----
  orderbook <- orderbook |>
    mutate(
      across(contains("bid_price"), ~ replace(., . < 0, NA)),
      across(contains("ask_price"), ~ replace(., . >= 999999, NA))
    )

  # Remove crossed order book observations ----
  orderbook <- orderbook |>
    filter(ask_price_1 > bid_price_1)

  # Merge transactions with unique timestamp ----
  trades <- orderbook |>
    filter(type == 4 | type == 5) |>
    select(ts:direction)

  trades <- inner_join(trades,
                       orderbook |>
                         group_by(ts) |>
                         filter(row_number() == 1) |>
                         ungroup() |>
                         transmute(ts,
                                   ask_price_1,
                                   bid_price_1,
                                   midquote = ask_price_1 / 2 + bid_price_1 / 2,
                                   lag_midquote = lag(midquote)
                         ),
                       by = "ts"
  )

  trades <- trades |>
    mutate(direction = case_when(
      type == 5 & m_price < lag_midquote ~ 1, # lobster convention: direction = 1 if executed against a limit buy order
      type == 5 & m_price > lag_midquote ~ -1,
      type == 4 ~ as.double(direction),
      TRUE ~ as.double(NA)
    ))

  # Aggregate transactions with size and volume-weighted price
  trade_aggregated <- trades |>
    group_by(ts) |>
    summarise(
      type = last(type),
      order_id = NA,
      m_price = sum(m_price * m_size) / sum(m_size),
      m_size = sum(m_size),
      direction = last(direction)
    )

  # Merge trades with the last observed order book snapshot
  trade_aggregated <- inner_join(trade_aggregated,
                                 orderbook |>
                                   select(ts,
                                          ask_price_1:last_col()) |>
                                   group_by(ts) |>
                                   filter(row_number() == n()),
                                 by = "ts"
  )

  orderbook <- orderbook |>
    filter(
      type != 4,
      type != 5
    ) |>
    bind_rows(trade_aggregated) |>
    arrange(ts) |>
    mutate(direction = if_else(type == 4 | type == 5,
                               direction,
                               as.double(NA)))

  return(orderbook)
}
```

You can process the raw data from before by simply calling.
```{r}
#| label: process-orderbook
orderbook <- process_orderbook(orderbook)
```

The function `compute_depth` below computes the number of traded shares within `bp` basis points of the associated best price, which can be used to measure liquidity. 
```{r}
#| label: define-compute-depth
compute_depth <- function(orderbook,
                          side = "bid",
                          bp = 0) {

  # Computes depth (in contracts) based on order book snapshots
  if (side == "bid") {
    value_bid <- (1 - bp / 10000) * orderbook |> select("bid_price_1")
    index_bid <- orderbook |>
      select(contains("bid_price")) |>
      mutate_all(function(x) {
        if_else(is.na(x),
                FALSE,
                x >= value_bid
        )
      })

    sum_vector <- (orderbook |> select(contains("bid_size")) * index_bid) |> rowSums(na.rm = TRUE)
  } else {
    value_ask <- (1 + bp / 10000) * orderbook |> select("ask_price_1")
    index_ask <- orderbook |>
      select(contains("ask_price")) |>
      mutate_all(function(x) {
        if_else(is.na(x),
                FALSE,
                x <= value_ask
        )
      })
    sum_vector <- (orderbook |> select(contains("ask_size")) * index_ask) |> rowSums(na.rm = TRUE)
  }
  return(sum_vector)
}
```

Next, I compute the midquote, bid-ask spread, aggregate volume, and depth (the amount of trade-able units in the order book):

- Midquote $q_t = (a_t + b_t)/2$ (where $a_t$ and $b_t$ denote the best bid and best ask)
- Spread $S_t= (a_t - b_t)$ (values below are computed in basis points relative to the concurrent midquote)
- Volume is the aggregate sum of traded units of the stock. I do not differentiate between hidden (`type==5`) and visible volume.
- Depth at the best level.

I aggregate the information into buckets of 5 minutes, using time-weighting for the order book variables.

```{r}
#| label: tab-orderbook-summary
#| tbl-cap: "Five-minute summaries of midquote, spread (bps), trade count, volume, and quoted depth for SAP, January 2022."
orderbook_summaries <- orderbook |>
  transmute(ts,
    midquote = ask_price_1 / 2 + bid_price_1 / 2,
    trading_volume = if_else(type == 4 | type == 5, m_price * m_size, as.double(NA)),
    depth0_bid = bid_size_1,
    depth0_ask = ask_size_1,
    spread = 10000 * (ask_price_1 - bid_price_1) / midquote
  ) |>
  mutate(
    ts_latency = as.numeric(lead(ts)) - as.numeric(ts), # time between messages
    ts_latency = if_else(is.na(ts_latency), 0, ts_latency), # 0 for first message
    ts_minute = lubridate::floor_date(ts, "5 minutes")
  ) |> # create 5 minute buckets (e.g. messages from 10:30:00 - 10:34:59.xx belong into 10:30:00)
  group_by(ts_minute) |>
  summarise(
    midquote = last(midquote),
    n_trades = sum(!is.na(trading_volume)),
    n = n(), # number of messages
    trading_volume = sum(trading_volume, na.rm = TRUE),
    trading_volume = if_else(is.na(trading_volume), as.double(0), trading_volume),
    depth0_bid = weighted.mean(depth0_bid, ts_latency),
    depth0_ask = weighted.mean(depth0_ask, ts_latency),
    spread = weighted.mean(spread, ts_latency)
  )

orderbook_summaries |>
  gt() |>
  cols_label(
    ts_minute      = "Time",
    midquote       = "Midquote (USD)",
    n_trades       = "Trades",
    n              = "Messages",
    trading_volume = "Volume (USD)",
    depth0_bid     = "Bid depth",
    depth0_ask     = "Ask depth",
    spread         = "Spread (bps)"
  ) |>
  tab_spanner(label = "Order book",       columns = c(midquote, spread, depth0_bid, depth0_ask)) |>
  tab_spanner(label = "Trading activity", columns = c(n_trades, n, trading_volume)) |>
  fmt_time(columns   = ts_minute,                      time_style = "hm") |>
  fmt_currency(columns = midquote,                     currency = "USD", decimals = 2) |>
  fmt_currency(columns = trading_volume,               currency = "USD", decimals = 0) |>
  fmt_number(columns   = c(depth0_bid, depth0_ask),    decimals = 0) |>
  fmt_number(columns   = spread,                       decimals = 2) |>
  fmt_integer(columns  = c(n_trades, n))
```

Finally, some visualization of the data at hand: The code below shows the dynamics of the traded prices (red points) and the quoted prices at the higher order book levels.

```{r}
#| label: fig-orderbook-dynamics
#| fig-cap: "Order book dynamics for SAP: quoted prices at each level (shaded by relative depth) with executed trades overlaid in red."
orderbook_trades <- orderbook |>
  filter(type==4|type==5) |>
  select(ts, m_price)

orderbook_quotes <- orderbook |>
    mutate(ts_new = floor_date(ts, "minute")) |>
    group_by(ts_new) |>
    summarise_all(dplyr::last) |>
    select(-ts_new) |>
    mutate(id = row_number()) |>
    select(ts, id, matches("bid|ask")) |>
    gather(level, price, -ts, -id) |>
    separate(level, into=c("side", "variable", "level"), sep="_") |>
    mutate(level = as.numeric(level))  |>
    pivot_wider(names_from = variable, values_from = price)

ggplot() +
  theme_bw() +
  geom_point(data = orderbook_quotes, aes(x=ts, y=price, color=as_factor(level), size = size/max(size)), alpha = 0.1)+
  geom_point(data = orderbook_trades, aes(x=ts, y=m_price), color='red') +
  labs(title="Orderbook Dynamics",
       y= "Price",
       x="") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position ="none") +
  scale_y_continuous()
```


## Python

No code available (yet)

:::

## Brownian Motion and realized volatility

In this exercise, you will analyze the empirical properties of the (geometric) Brownian motion based on a simulation study. Such simulation studies can also be helpful for various applications in continuous time finance, e.g., option pricing.

Finance theory suggests that prices must be so-called semimartingales. The most commonly used semimartingale is the Geometric Brownian Motion (GBM), where we assume that the logarithm of the stock price $S_t$, $X_t = \log(S_t)$ follows the following dynamic
$$
X_t = X_0 + \mu t + \sigma W_t
$$
where $\mu$ and $\sigma$ are constants and $W_t$ is a Brownian motion.

**Exercises:**

- Write a function that simulates (log) stock prices according to the model above. Recall that the Brownian motion has independent increments which are normally distributed. The function should generate a path of the log stock price with a pre-defined number of observations (granularity), drift term, and volatility. Then, plot one sample path of the implied (log) stock price.
- Use the function to simulate a large number of geometric Brownian motions. Illustrate the quadratic variation of the process. 
- Compute the realized volatility for each simulated process and illustrate the sampling distribution. 
- Finally, consider the infill asymptotics. Show the effect of sampling once per day, once per hour, once per minute, and so on on the variance of the realized volatility. 

**Solutions:**

::: {.panel-tabset group="language"}
## R

For this exercise, I use some functionalities from the `xts` and `highfrequency` packages, particularly for aggregation at different frequencies.
```{r}
#| label: load-libraries-bm
library(dplyr)
library(tidyr)
library(xts)
library(highfrequency)
library(purrr)
```

**Note:** When loading `xts` and the `tidyverse` jointly, you may run into serious issues if you do not consider naming conflicts! Especially the convenient functions `last` and `first` from `dplyr` may be affected. For that purpose, I consistently use `dplyr::first` and `dplyr::last` in this chapter.

First, the function below is a rather general version of a simulated generalized Brownian motion. Note that there is no immediate need to assign trading date and hour; I implement this because it will provide output closer to actual data, and therefore, parts of my code can be recycled for future projects. The function can be called with a specific seed to replicate the results. The relevant parameters are the initial values `start` and the geometric Brownian motion `mu` and `sigma` parameters. The input `steps` defines the simulated Brownian motion's granularity (sampling frequency).

```{r}
#| label: define-simulate-gbm
simulate_geom_brownian_motion <- function(seed = 3010,
                                          steps = 1000,
                                          mu = 0,
                                          sigma = 1,
                                          start = 0){
  set.seed(seed)
  date <- Sys.Date() + days(seed)
  hour(date) <- 9
  minute(date) <- 30
  date_end <- date
  hour(date_end) <- 16
  minute(date_end) <- 0

  tibble(step = 1:(steps+1),
         time =   seq(from = date,
                      to = date_end,
                      length.out = steps + 1),
         price = cumsum(c(start, rnorm(steps, mean = mu/steps, sd = sigma/sqrt(steps)))))
}
```

Note the last line of the function doing most of the work: We exploit that increments of a Brownian motion are independent and normally distributed. The entire path is nothing but the cumulative sum of the increments.

Next, I simulate the stock price with 1000 intraday observations. Recall that the geometric Brownian motion implies dynamics for the log of the stock price, which is why I adjust the plotted value in the $y$-axis.
```{r}
#| label: fig-gbm-single-path
#| fig-cap: "A single simulated intraday path of the geometric Brownian motion (1000 steps)."
tab <- simulate_geom_brownian_motion(start = log(18))
tab |>
  ggplot(aes(x=time, y = exp(price))) +
  geom_line() + labs(x="", y = "Stock price")
```

Next, we scale the simulation up. The code below simulates 250 paths of the geometric Brownian motion, all with the same starting value. The function illustrates the sampled (log) price paths.
```{r}
#| label: fig-gbm-multiple-paths
#| fig-cap: "Two hundred and fifty simulated log price paths of the geometric Brownian motion ($\\sigma = 2.4$, 1000 intraday steps each)."
true_sigma <- 2.4
steps <- 1000

simulation <- tibble(iteration = 1:250)
simulation <- simulation |>
  mutate(values = map(iteration, simulate_geom_brownian_motion, sigma = true_sigma, steps = steps)) |>
  unnest(values)

simulation |>
  ggplot(aes(x = step, y = price, group = iteration)) +
  geom_line(alpha = 0.3) +
  labs(x = "Iteration", y = "(log) price")
```

Next, we investigate some of the properties of the geometric Brownian motion. First, by definition,
the sample variation of $X_t = \int\limits_0^t\mu dt + \int\limits_0^t\sigma dW_t$ is $\int\limits_0^t\sigma^2 dt = \sigma^2t$. The figure below illustrates the sample variation over time and shows that the standard deviation across all simulated (log) prices increases linearly in $t$ as expected. For the simulation, I used $\sigma = 2.4$ (`true_sigma`), which corresponds to the slope of the line (adjusted by the number of observations).

```{r}
#| label: fig-gbm-sample-variation
#| fig-cap: "Sample variation of simulated GBM log prices over time; the straight line shows the theoretical $\\sigma^2 t$."
simulation |>
  group_by(step) |>
  summarise(sd = sd(price)) |>
  ggplot(aes(x = step, y = sd^2)) +
  geom_line() +
  geom_abline(aes(slope = true_sigma^2/steps, intercept = 0)) +
  labs(x = "Iteration", y = "Sample variation")
```

What about the realized volatility? As defined in the lecture, the realized variance $$RV = \sum\limits_{i=0}^n \Delta X_t^2$$ converges to the integrated variance (in the absence of jumps). The bar plot below illustrates the sampling distribution of the realized volatility, which, as expected, centers around `true_sigma`, the integrated variance of the simulated geometric Brownian motion. You can change the value `true_sigma` above and analyze the effect.
```{r}
#| label: fig-rv-distribution
#| fig-cap: "Sampling distribution of the realized variance across 250 simulated GBM paths; the dotted vertical line marks the true integrated variance."
simulation |>
  group_by(iteration) |>
  summarise(rv = sum(((price - lag(price))^2), na.rm = TRUE)) |>
  ggplot(aes(x = rv)) +
  geom_histogram(bins = 100) +
  labs(x = "Realized Volatility") +
  geom_vline(aes(xintercept = true_sigma^2), color = "red", linetype = "dotted")
```

Finally, some more analysis regarding the infill asymptotics. As should be clear by now, sampling at the highest frequency possible will ensure that the realized volatility recovers the integrated variance of the (log) price process. In other words, we can measure the variance. The following lines of code illustrate the effect: Whereas we simulated our process at a high frequency, we now simulate sub-samples at different frequencies, ranging from every second to once per day. The figure then illustrates the daily realized volatilities relative to the integrated variance of the process. Not surprisingly, the dispersion around the true parameter decreases substantially with higher sampling frequency.

Note that the code below illustrates how to work with the `xts` package and switch between the `tidyverse` and the `high-frequency` data format requirements. Of course, the procedure could be more convenient, but it is worth exploiting the functionalities of both worlds instead of recording everything in a tidy manner.

```{r}
#| label: fig-infill-asymptotics
#| fig-cap: "Sampling distributions of the realized volatility at six different sampling frequencies, illustrating infill asymptotics; the dotted line marks the true volatility $\\sigma$."
#| cache: true
infill_asymptotics <- function(tab, sampling_frequency = 5){
  tab_xts<- as.xts(tab |> select(price),
                 order.by = tab |> pull(time))
tab_xts <- aggregatePrice(tab_xts,
                                  alignBy = "minutes", alignPeriod = sampling_frequency,
                                  tz = "GMT",  marketOpen = "09:30:00")

  tab_xts |>
    fortify.zoo() |>
    as_tibble() |>
    rename("ts" = Index,
           "price" = PRICE) |>
    summarise(rv = sum((price -lag(price))^2, na.rm = TRUE))|>
    as.numeric()
}

rvs <- simulation |>
  group_by(iteration) |>
  nest(data = c(step, time, price)) |>
  mutate("1 Second" = map_dbl(data, infill_asymptotics, 1/60),
         "1 Minute" = map_dbl(data, infill_asymptotics, 1),
         "5 Minutes" = map_dbl(data, infill_asymptotics, 5),
         "15 Minutes" = map_dbl(data, infill_asymptotics, 15),
         "1 hour" = map_dbl(data, infill_asymptotics, 60),
         "1 day" = map_dbl(data, infill_asymptotics, 6.5 * 60))

rvs |>
    select(-data) |>
    ungroup() |>
    pivot_longer(-iteration,
                 names_to = "Sampling Frequency",
                 values_to = "rv") |>
    mutate(`Sampling Frequency` = as.factor(`Sampling Frequency`),
           `Sampling Frequency` = fct_reorder(`Sampling Frequency`, rv, sd)) |>
    ggplot(aes(x = sqrt(rv),
               fill = `Sampling Frequency`)) +
    geom_density() +
    geom_vline(aes(xintercept = true_sigma), linetype ="dotted") +
    facet_wrap(~`Sampling Frequency`, scales = "free_y") +
    labs(x = "(Log) Realized Volatility", y = "")
```

## Python

:::